Large-scale feature selection with Gaussian mixture models for remote sensing images

### 2016-10-14 paper.tex
relecture complete

% \tableofcontents
% \clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This article presents a forward feature selection algorithm based on Gaussian mixture model (GMM) classifiers. The algorithm selects iteratively features that maximize a criterion function which can be either a classification rate or a measure of divergence. Several variations of this algorithm are explored by changing the criterion function and also by testing a floating forward variation allowing backward step to discard already selected features.

An important effort is made in exploiting GMM properties to implement a fast algorithm. In particular, update rules of the GMM model are used to compute the criterion function with various sets of features. The result is a C++ remote module for the remote sensing processing toolbox Orfeo (OTB) developed by CNES.

Finally, the method is compared to standard classifiers using two different data-sets, one of hyperspectral images with a lot of spectral variables and one with heterogeneous spatial features. The results validate the fact that the method performs well in terms of processing time and classification accuracy in comparison to the standard classifiers available in OTB.\\
\end{abstract}

@@

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched successfully recently and 2 Terabytes of data will be released every days~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has released thousands of images. This high volume of Earth observation data provides an accurate information of land state and functioning, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

@@

A method  to reduce  both statistical and  computational issues  is to
perform a reduction of the dimension. In fact, with the \emph{curse of
  dimensionality}     comes     the      \emph{blessing     of     the
  dimensionality}~\cite{bouveyron2014model}:  high   dimensional  data
space   exhibits  interesting   properties   for  the   classification
purpose.  In  particular,  it  is   possible  to  get  a  parsimonious
representation  of  the  data  while  maintaining  or  increasing  the
classification  accuracy~\cite{jimenez1998supervised}.  For  instance,
in land-cover  classification, given  a set  of spatial,  temporal and
spectral features, it is possible to  extract those which are the most
discriminant      for      the     purpose      of      classification
\cite{fassnacht2014comparison}.   In  hyperspectral   data,  from  the
hundreds of available spectral channels,  it is possible to reduce the
number of channels  to make the processing more efficient  in terms of
statistical complexity  and computational load. In  short, by reducing
the dimension better classification results is expected with a reduced
computational load.

@@

There  are  two  main  strategies  to  reduce  dimension~\cite[Chapter
1]{Guyon:2006:FEF:1208773}: features extraction and feature selection.
Feature extraction means reformulate  and summarize the information by
creating new features in combining  the existing ones, it is sometimes
referred to  \emph{feature construction}.   Linear combination  of the
initial features  can be extracted using  Principal Component Analysis
(PCA)~\cite{jimenez1998supervised}  or Independent  Component Analysis
\cite{villa2011hyperspectral}.  Supervised extraction  method has also
been investigated  such as Fischer discriminant  analysis and decision
boundary feature  extraction~\cite{book:landgrebe}.  To  the contrary,
features selection  extracts a subset of  existing features identified
as the  most relevant  by a  given criterion.   It has  the additional
advantage to be much more easily understandable for the end-user.

@@

The third class corresponds to the \emph{embedded method}. They do not separate the features selection process from the learning algorithm and allow interaction between the two processes. A popular method embedded method is the \emph{Random Forest}. Embedded methods also exist for other models, \emph{e.g.} SVM \cite{guyon2002gene,weston2003use,tuia2015multiclass}.

@@

The remaining of the article is organized as follows. Section~\ref{sec:gmm-hd} presents GMM classifiers and the problems related to high-dimensional feature space. The features selection methods are detailed in Section~\ref{sec:selection}. Then an efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. Experimental results on two real high dimensional data set are given Section~\ref{sec:test}. Conclusion and perspectives conclude the paper in Section~\ref{sec:conclusion}.

\section{Gaussian Mixture Models in high dimensional space}
\label{sec:gmm-hd}

@@

    %% Question: tu suggérais de reprendre ici "Ensuite, il y a la réduction de dimension données des refs (voir intro du papier JSTARS) et dans la réduction  de dimension, il  y a  la sélection de features." Le soucis que je vois, c'est que je détaille déjà dans l'intro les différentes méthodes de sélection/extraction et je trouve que ça devient pas très clair si je reprends tout en me basant sur le papier JSTARS. J'aurais dit avt "il y a trois types de sélection: filtre, wrapper, embedded blablabla" et ensuite ça me fait reprendre ici une autre distinction "supervisé/non-supervisé" si je me base sur jstars qui est un peu redondante il me semble. Qu'en dis-tu ?
    Second option is to reduce the dimension. Features extraction/selection methods have been developed in order to reduce the dimension with various approaches described in Section~\ref{sec:intro}. In this study, this latter option is explored and a feature selection method named sequential forward features selection is presented after an overview of the selection algorithms.

@@

The feature selection method proposed in this work is a wrapper method associated to GMM models. Two elements are needed to set up a wrapper method:
\begin{enumerate}
\item A function that  ranks the features according to some good classification criteria,
\item A search strategy to optimize the function.
\end{enumerate}

@@

        The \emph{Jeffries–Matusita distance} is a measure based on the Bhattacharyya distance. It saturates if the separability between the two distribution increases~\cite{bruzzone2009novel}. The JM distance is defined as
        \begin{equation}\label{eq:jm}
            {JM}_{cc^\prime} = \sqrt{ 2 \{1 - \text{exp}(-B_{cc^\prime})\} }.
        \end{equation}

@@

        \subsection{Selection method}
        \label{sec:selection:method}
        Two  sequential  search  algorithms were  implemented  in  this
        work~\cite{Guyon:2006:FEF:1208773}:  the   sequential  forward
        selection and the sequential  floating forward.  The later one
        is  an   extension  of   the  former.  Both   select  features
        iteratively.

@@

An efficient implementation of the criterion optimization is proposed.
It based on  the symmetry property of the covariance  matrix and block
inverse  formula~\cite{IMM2012-03274}. In  the following  it is  shown
that the criterion  can be split into  two part: one that  needs to be
computed for each  tested variable, and one that needs  to be computed
only once per  selection step. For the  cross-validation part, updates
rules are given to derive sub-models  without the necessity to learn a
GMM models for each fold.

@@

      \begin{prop}[Kullback-Leibler divergence~(\ref{eq:skl})]
        \begin{align}
        \label{eq:skl-update}
            & SKL_{cc^\prime} = \frac{1}{2} \biggl\{ \underbrace{\text{\normalfont Tr} \left((\boldsymbol{\Sigma}_c^{(k)})^{-1} \boldsymbol{\Sigma}_{c^\prime}^{(k)} + (\boldsymbol{\Sigma}_{c^\prime}^{(k)})^{-1} \boldsymbol{\Sigma}_c^{(k)}\right)}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}_c^t & \frac{1}{\alpha_c} \end{array}\right] (\boldsymbol{\mu}_c^{(k)} - \boldsymbol{\mu}_{c^\prime}^{(k)}) )^2}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}_{c^\prime}^t & \frac{1}{\alpha_{c^\prime}} \end{array}\right] (\boldsymbol{\mu}_c^{(k)} - \boldsymbol{\mu}_{c^\prime}^{(k)}) )^2 - 2k}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{(\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1}(\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber\\
          & + \underbrace{(\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})^t(\boldsymbol{\Sigma}_{c^\prime}^{(k-1)})^{-1} (\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})}_{\substack{\text{computed once per selection step}}} \biggr\}.
        \end{align}
        with $(\boldsymbol{\Sigma}_c^{(k)})^{-1}$ computed with Proposition~\ref{eq:update-inv};
      \end{prop}

@@

    Same reasoning applied  to the term $\alpha$ in  the update rules:
    it       is       also        thresholded       to       EPS\_FLT.
    Algorithm~\ref{alg:sffs-update}    details   when    computational
    stability is enforced (lines 6, 10, 20 and 24).

@@

    Following  OTB framework,  two  applications  are available.   The
    first,  called  \texttt{otbcli\_TrainGMMSelectionApp}, performs  a
    feature selection algorithm  (SFS or SFFS) with  the one criterion
    given      from~\ref{tab:crit}.       The      second,      called
    \texttt{otbcli\_PredictGMMApp},   generates   the  thematic   maps
    according to the  learn model. The only  technical limitation that
    the  training set  must  fit  in the  RAM  of  the computer.   The
    classification step is streamed and there is no limitation in term
    of image  size. The Figure~\ref{fig:otb:ffs} shows  a code excerpt
    to run the application.

@@

    In order to increase the dimensionality of the data, the following features are computed using the RGBIR images similarly to \cite{tuia2015multiclass}:
    \begin{itemize}
        \item Fifteen Radiometric indexes: NDVI, TNDVI, RVI, SAVI, TSAVI, MSAVI, MSAVI2, GEMI, IPVI, NDWI2, NDTI, RI, CI, BI, BI2~\cite{otb}.
        \item Morphological profile build on each band with a disk of radius 5, 9, 13, 17, 21, 25, 29, 33, 37 and 41 (80 features)~\cite{fauvel2013advances};
        \item Attribute profile build on each band with area as attribute and 1000, 2000, 5000, 10000 and 15000 as thresholds (40 features)~\cite{dalla2010morphological}.
        \item Attribute profile build on each band with diagonal of bounding box as attribute and 100, 200, 500, 1000 and 20000 as thresholds (40 features)~\cite{dalla2010morphological}.
        \item Textural features for  each channel with neighborhood of
          19x19 pixels:  mean, standard  deviation, range  and entropy
          (16 features)~\cite{otb}.
    \end{itemize}
    The normalized DSM and the raw  RGBIR image are added to these 191
    features and  then then  stacked to  create a  new image  with 196
    bands. %  The  training is  made with  tile 5\_11  and test  with two
    % tiles: 5\_12 which very similar with  tile 5\_11 and 3\_10 which a
    % further area a bit less urban and more industrial.

>>>>

% \tableofcontents
% \clearpage
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This article presents a forward feature selection algorithm based on Gaussian mixture model (GMM) classifiers. The algorithm selects iteratively features that maximize a criterion function which can be either a classification rate or a measure of divergence. Several variations of this algorithm are explored by changing the criterion function and also by testing a floating forward variation allowing backward step to discard already selected features.
An important effort is made in exploiting GMM properties to implement a fast algorithm. In particular, update rules of the GMM model are used to compute the criterion function with various sets of features. The result is a C++ remote module for the remote sensing processing toolbox Orfeo (OTB) developed by CNES.
Finally, the method is compared to standard classifiers using two different data-sets, one of hyperspectral images with a lot of spectral variables and one with heterogeneous spatial features. The results validate the fact that the method performs well in terms of processing time and classification accuracy in comparison to the standard classifiers available in OTB.\\
\end{abstract}

@@

\IEEEPARstart{W}{ith} the increasing number of remote sensing missions, the quantity of available Earth observation data for a given landscape becomes larger and larger. Satellite missions produce a huge amount of data on a regularly (daily) basis. From 2018, the EnMAP (Environmental Mapping and Analysis Program) satellites managed by the German space agency will produce images with 244 spectral bands, a spatial resolution of 30x30m per pixel and with a frequency of revisit of 4 days\cite{Müller09enmap}. The Hyperspectral Infrared Imager (HyspIRI) of NASA will also deliver images with 212 spectral bands. Additionally to hyperspectral data, the amount of available hypertemporal data increases a lot. For instance, the European satellites Sentinel-2 were launched successfully recently and 2 Terabytes of data will be released every days~\cite{drusch2012sentinel}. The Landsat open archive (\url{http://landsat.usgs.gov/products_data_at_no_charge.php}) has released thousands of images. This high volume of Earth observation data provides an accurate information of land state and functioning, and helps to improve the understanding of the planet~\cite{rs1010001}. However, processing such data is more and more challenging because of statistical and computational issues.

@@

A method  to reduce  both statistical and  computational issues  is to
perform a reduction of the dimension. In fact, with the \emph{curse of
  dimensionality}     comes     the      \emph{blessing     of     the
  dimensionality}~\cite{bouveyron2014model}:  high   dimensional  data
space   exhibits  interesting   properties   for  the   classification
purpose.  In  particular,  it  is   possible  to  get  a  parsimonious
representation  of  the  data  while  maintaining  or  increasing  the
classification  accuracy~\cite{jimenez1998supervised}.  For  instance,
in land-cover  classification, given  a set  of spatial,  temporal and
spectral features, it is possible to  extract those which are the most
discriminant      for      the     purpose      of      classification
\cite{fassnacht2014comparison}.   In  hyperspectral   data,  from  the
hundreds of available spectral channels,  it is possible to reduce the
number of channels  to make the processing more efficient  in terms of
statistical complexity  and computational load. In  short, by reducing
the dimension better, classification results is expected with a reduced
computational load.

@@

There  are  two  main  strategies  to  reduce  dimension~\cite[Chapter
1]{Guyon:2006:FEF:1208773}: features extraction and feature selection.
Feature extraction means reformulate  and summarize the information by
creating new features in combining  the existing ones, it is sometimes
referred to as \emph{feature construction}.   Linear combination  of the
initial features  can be extracted using  Principal Component Analysis
(PCA)~\cite{jimenez1998supervised}  or Independent  Component Analysis
\cite{villa2011hyperspectral}.  Supervised extraction  method has also
been investigated  such as Fischer discriminant  analysis and decision
boundary feature  extraction~\cite{book:landgrebe}.  To  the contrary,
features selection  extracts a subset of  existing features identified
as the  most relevant  by a  given criterion.   It has  the additional
advantage to be much more easily understandable for the end-user.

@@

The third class corresponds to the \emph{embedded methods}. They do not separate the features selection process from the learning algorithm and allow interaction between the two processes. A popular method embedded method is the \emph{Random Forest}. Embedded methods also exist for other models, \emph{e.g.} SVM \cite{guyon2002gene,weston2003use,tuia2015multiclass}.

@@

The remaining of the article is organized as follows. Section~\ref{sec:gmm-hd} presents GMM classifiers and the problems related to high-dimensional feature spaces. The features selection methods are detailed in Section~\ref{sec:selection}. Then an efficient implementation of the proposed selection method is presented in Section~\ref{sec:implementation}. Experimental results on two real high dimensional data-set are given Section~\ref{sec:test}. Conclusion and perspectives conclude the paper in Section~\ref{sec:conclusion}.

\section{Gaussian Mixture Models in high dimensional spaces}
\label{sec:gmm-hd}

@@

    Second option is to reduce the dimension. Features extraction/selection methods have been developed in order to reduce the dimension with various approaches described in Section~\ref{sec:intro}. In this study, this latter option is explored and a feature selection method named sequential forward features selection is presented.

@@

The feature selection method proposed in this work is a wrapper method associated to GMM models. Two elements are needed to set up a wrapper method:
\begin{enumerate}
\item A function that ranks the features according to some good classification or class separability criteria,
\item A search strategy to optimize the function.
\end{enumerate}

@@

        The \emph{Jeffries-Matusita distance} is a measure based on the Bhattacharyya distance. It saturates if the separability between the two distributions increases~\cite{bruzzone2009novel}. The JM distance is defined as
        \begin{equation}\label{eq:jm}
            {JM}_{cc^\prime} = \sqrt{ 2 \{1 - \text{exp}(-B_{cc^\prime})\} }.
        \end{equation}

@@

        \subsection{Selection method}
        \label{sec:selection:method}
        Two  sequential  search  algorithms have been  implemented  in  this
        work~\cite{Guyon:2006:FEF:1208773}:  the   sequential  forward
        selection and the sequential  floating forward.  The later one
        is  an   extension  of   the  former.  Both   select  features
        iteratively.

@@

An efficient implementation of the criterion optimization is proposed.
It is based on  the symmetry property of the covariance  matrix and block
inverse  formula~\cite{IMM2012-03274}. In  the following  it is  shown
that the criterion  can be split into  two parts: one that  needs to be
computed for each  tested variable, and one that needs  to be computed
only once per  selection step. For the  cross-validation part, updates
rules are given to derive sub-models  without the necessity to learn a
GMM models for each fold.

@@

      \begin{prop}[Kullback-Leibler divergence~(\ref{eq:skl})]
        \begin{align}
        \label{eq:skl-update}
            & SKL_{cc^\prime} = \frac{1}{2} \biggl\{ \underbrace{\text{\normalfont Tr} \left((\boldsymbol{\Sigma}_c^{(k)})^{-1} \boldsymbol{\Sigma}_{c^\prime}^{(k)} + (\boldsymbol{\Sigma}_{c^\prime}^{(k)})^{-1} \boldsymbol{\Sigma}_c^{(k)}\right)}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}_c^t & \frac{1}{\alpha_c} \end{array}\right] (\boldsymbol{\mu}_c^{(k)} - \boldsymbol{\mu}_{c^\prime}^{(k)}) )^2}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}_{c^\prime}^t & \frac{1}{\alpha_{c^\prime}} \end{array}\right] (\boldsymbol{\mu}_c^{(k)} - \boldsymbol{\mu}_{c^\prime}^{(k)}) )^2 - 2k}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{(\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})^t (\boldsymbol{\Sigma}_c^{(k-1)})^{-1}(\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber\\
          & + \underbrace{(\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})^t(\boldsymbol{\Sigma}_{c^\prime}^{(k-1)})^{-1} (\boldsymbol{\mu}_c^{(k-1)} - \boldsymbol{\mu}_{c^\prime}^{(k-1)})}_{\substack{\text{computed once per selection step}}} \biggr\},
        \end{align}
      \end{prop}
      \noindent with $(\boldsymbol{\Sigma}_c^{(k)})^{-1}$ computed with Proposition~\ref{eq:update-inv}.

@@

    Same reasoning applied  to the term $\alpha$ in  the update rules:
    it       is       also        thresholded       to       EPS\_FLT.
    Algorithm~\ref{alg:sffs-update}    details   when    computational
    stability is enforced (lines 7, 13, 25 and 31).

@@

    Following  OTB framework,  two  applications  are available.   The
    first,  called  \texttt{otbcli\_TrainGMMSelectionApp}, performs  a
    feature selection algorithm  (SFS or SFFS) with  the one criterion
    given      from Table~\ref{tab:crit}.       The      second,      called
    \texttt{otbcli\_PredictGMMApp},   generates   the  thematic   maps
    according to the  learn model. The only  technical limitation that
    the  training set  must  fit  in the  RAM  of  the computer.   The
    classification step is streamed and there is no limitation in term
    of image  size. The Figure~\ref{fig:otb:ffs} shows  a code excerpt
    to run the application.

@@

    In order to increase the dimensionality of the data, the following features are computed using the RGBIR images similarly to \cite{tuia2015multiclass}:
    \begin{itemize}
        \item Fifteen Radiometric indexes: NDVI, TNDVI, RVI, SAVI, TSAVI, MSAVI, MSAVI2, GEMI, IPVI, NDWI2, NDTI, RI, CI, BI, BI2~\cite{otb}.
        \item Morphological profile build on each band with a disk of radius 5, 9, 13, 17, 21, 25, 29, 33, 37 and 41 (80 features)~\cite{fauvel2013advances};
        \item Attribute profile build on each band with area as attribute and 1000, 2000, 5000, 10000 and 15000 as thresholds (40 features)~\cite{dalla2010morphological}.
        \item Attribute profile build on each band with diagonal of bounding box as attribute and 100, 200, 500, 1000 and 20000 as thresholds (40 features)~\cite{dalla2010morphological}.
        \item Textural features for  each channel with neighborhood of
          19x19 pixels:  mean, standard  deviation, range  and entropy
          (16 features)~\cite{otb}.
    \end{itemize}
    The normalized DSM and the raw  RGBIR image are added to these 191
    features and  then stacked to  create a  new image  with 196
    bands. %  The  training is  made with  tile 5\_11  and test  with two
    % tiles: 5\_12 which very similar with  tile 5\_11 and 3\_10 which a
    % further area a bit less urban and more industrial.
